The Key to a Data Parallel Compiler

### 2016-04-07 2016-spring-array/2016-array-key.tex
Updated with some comments by Jason Hemann

That is, the sample code given indented by 6 spaces, followed by the value 
of the expression without indent. All expressions are evaluated right to 
left and all function application is unified and infix; that is, a function 
application may apply a function (such as \verb;+;) to one or two arguments, 
called monadic and dyadic application respectively,
which must appear on the right and (optionally) on the left of the function 
name, as in the above example. Tables \ref{tab:scalarprims} to 
\ref{tab:conjunctions}
provide a listing of all of the primitives used in the Co-dfns compiler, 
a selection of which are used here.

@@

\begin{table*}
\centering
\begin{tabular}{l p{2.8in} l l l}
\toprule
Pass & Description & Core Patterns & Coordinate? \\
\midrule
Record Node Coordinates & Adds a new field to each node containing that node's 
 node coordinate. & Outer Product/Scan & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & Yes \\
Fold Constants & Converts constant expressions to literals & Amend & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No \\
Lift Type-checking & Infers some type information at compile time &
 Power Limit/Rank & No \\
Allocate Value Slots & Does a form of frame allocation for variables &
 Key & No \\
Anchor Variables & Resolves lexically scoped variables &
 Key & Yes \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
 Key & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{tab:passes}
\end{table*}

@@

\section{Other Passes}

Table \ref{tab:passes} contains a selection of passes from the
Co-dfns compiler and breaks down their features based on how they
traverse the tree and whether or not they use the Key operator or
Node Coordinates. They are oriented with front-end passes higher in
the table and passes further down the compiler chain at the bottom
of the table. These passes give a good idea of how the Key operator
plays a part in the complete compiler as well as how Node Coordinates
help to deal with complex sub-tree selections.

In short, the goal of the structure and design of the compiler passes
given in table \ref{tab:passes} is to utilize Node Coordinates to
remove as quickly as possible the nested structure of the AST into
a flatter form that does not require complex analysis to traverse.

\section{The Co-dfns Compiler}

The Co-dfns compiler is an experimental commercial compiler funded by 
Dyalog, Ltd. \cite{hsu2014co,hsu2015accelerating}
It is intended to provide increased scaling and performance
for APL programmers using the dfns syntax (lexically scoped, functionally 
oriented APL). These techniques form one of the foundational elements to 
the architecture of the compiler, which is built in the style of a Nanopass 
compiler, with very small passes chained together through composition, with 
the added requirements that each compiler pass must be fully data parallel.

@@

@@

Work remains on expanding the features and capabilities of the compiler, 
to demonstrate the scalability of these techniques on more sophisticated 
optimizations. The current compiler represents a non-trivial instance of 
the successful application of these techniques, but it focuses on an untyped, 
functionally oriented language. The type inference in the current version 
of the compiler is somewhat naive. More work is required to expand the 
type inferencer to include more sophisticated type inference. Likewise, 
the compiler does not currently handle error handling such as signals or 
exceptions.

Further work remains to implement these techniques on an imperative and 
object-oriented language instead of a functional one. It would also be 
educational to implement a version for a lazy language.

The performance of these techniques has not been a problem in practice, but 
we do not have a large scale understanding of the performance characteristics 
of these techniques in real life. Further implementation and analysis is 
required to understand what sort of compiler optimizations and analyses are 
necessary to compile the Co-dfns compiler efficiently for the GPU and CPU. 
We current lack a good comparison against existing techniques for the same 
language, since the current compiler is relatively unique in that no other 
compilers to our knowledge compile the same language to the same targets 
with the same sorts of optimizations. We intend to implement other paradigms 
and other languages which do have existing compiler implementations to compare 
relative performance on multiple platforms.

>>>>

That is, we indent the expression by 6 spaces, followed by the value
of the expression without indent. All expressions are evaluated right to 
left and all function application is infix; that is, a function 
application may apply a function (such as \verb;+;) to one or two arguments, 
called monadic and dyadic application respectively,
which must appear on the right and (optionally) on the left of the function 
name, as in the above example. Tables \ref{tab:scalarprims} to 
\ref{tab:conjunctions}
provide a listing of all of the primitives used in the Co-dfns compiler, 
a selection of which are used here.

@@

\begin{table*}
\centering
\begin{tabular}{l p{2.8in} l l l}
\toprule
Pass & Description & Core Traversal Pattern & Uses coordinates? \\
\midrule
Record Node Coordinates & Adds a new field to each node containing that node's 
 node coordinate. & Outer Product/Scan & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & Yes \\
Fold Constants & Converts constant expressions to literals & Amend & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No \\
Lift Type-checking & Infers some type information at compile time &
 Power Limit/Rank & No \\
Allocate Value Slots & Does a form of frame allocation for variables &
 Key & No \\
Anchor Variables & Resolves lexically scoped variables &
 Key & Yes \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
 Key & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{tab:passes}
\end{table*}

@@

\section{The Co-dfns Compiler and Other Passes}

The Co-dfns compiler is an experimental commercial compiler funded by 
Dyalog, Ltd. \cite{hsu2014co,hsu2015accelerating}
It is intended to provide increased scaling and performance
for APL programmers using the dfns syntax (lexically scoped, functionally 
oriented APL). These techniques form one of the foundational elements to 
the architecture of the compiler, which is built in the style of a Nanopass 
\cite{keep2013nanopass}
compiler, with very small passes chained together through composition, with 
the added requirements that each compiler pass must be fully data parallel.

@@

Table \ref{tab:passes} contains a selection of passes from the
Co-dfns compiler and breaks down their features based on how they
traverse the tree and whether or not they use the Key operator or
Node Coordinates.
The traversal pattern indicates the primary operator used to traverse the AST 
and perform the transformation, most of which are primitives; the Amend pattern 
is a ``replacement'' pattern where specific nodes are replaced in the AST with 
other nodes, and can be thought of as a slightly modified filter and map.
The table orders the passes with front-end passes higher in
the table and passes further down the compiler chain at the bottom
of the table. These passes give a good idea of how the Key operator
plays a part in the complete compiler as well as how Node Coordinates
help to deal with complex sub-tree selections.

In short, the goal of the structure and design of the compiler passes
given in Table \ref{tab:passes} is to utilize Node Coordinates to
remove as quickly as possible the nested structure of the AST into
a flatter form that does not require complex analysis to traverse.

@@

The current compiler represents a non-trivial instance of
the successful application of these techniques, but it focuses on an untyped, 
functionally oriented language. The type inference in the current version 
of the compiler is somewhat naive. More work is required to expand the 
type inferencer to include more sophisticated type inference. 
The compiler does not currently handle errors such as signals or 
exceptions, nor does it implement some more sophisticated compiler optimizations.
Further work remains to implement these techniques on an imperative and 
object-oriented language instead of a functional one. It would also be 
educational to implement a version for a lazy language.

We do not have a large scale understanding of the performance characteristics 
of these techniques in real life outside of daily use of the compiler.
Further analysis is
required to understand what sort of compiler optimizations and analyses are 
necessary to compile the Co-dfns compiler efficiently for the GPU and CPU. 
We currently lack a good comparison against existing techniques for the same 
language, since the current compiler is relatively unique in that no other 
compilers to our knowledge compile the same language to the same targets 
with the same sorts of optimizations. We intend to implement other paradigms 
and other languages which do have existing compiler implementations to compare 
relative performance on multiple platforms.
